RNeXML: Parsing and Serializing the Next Generation of Phyloinformatic Data in R

### 2014-06-16 manuscripts/manuscript.Rmd
edits, more work on intro, fixed some citations

---
layout: preprint
#layout: review, 11pt
title: "RNeXML: Parsing and Serializing the Next Generation of Phyloinformatic Data in R"
author: 
  - name: Carl Boettiger
    affiliation: cstar
    email: cboettig@gmail.com
    footnote: Corresponding author
  - name: Hilmar Lapp
    affiliation: NESCent
  - name: Scott Chamberlain
    affiliation: SimonFraser
  - name: Rutger Vos
    affiliation: NBC
address: 
  - code: cstar
    address: | 
      Center for Stock Assessment Research, 
      Department of Applied Math and Statistics, 
      University of California, Mail Stop SOE-2,
      Santa Cruz, CA 95064, USA
  - code: NESCent
    address: | 
      National Evolutionary Synthesis Center, Durham, NC, USA
  - code: SimonFraser
    address: |
      Department of Biology, Simon Fraser University, Burnaby, Canada
  - code: NBC
    address: | 
      Naturalis Biodiversity Center, Leiden, the Netherlands
abstract: | 
      1. NeXML promises to be the next generation of phylogenetic
      informatics data and metadata exchange formats, offering substantial
      improvements in reliability, extensibility, and richness over
      earlier standards such as NEXUS. In this paper we present the RNeXML
      package which provides a bridge between this rich and reliable
      data format and the extensive library of phylogenetic tools already
      available in R, with a particular emphasis on comparative methods.

      2. In this paper we present the RNeXML package, which provides
      a bridge between this rich and reliable data format and the
      extensive library of phylogenetic tools already available in R,
      with a particular emphasis on comparative methods. 
      RNeXML is much richer than a basic parser: here we illustrate
      how the package facilitates adding and manipulating semantic 
      annotations.

@@

      4. RNeXML is perfectly suited for users to preserve and maintain their
      personal archives of phylogenetic and comparative trait data in consistent,
      query-able, and cross-platform format, or leverage the rapidly growing 
      public community archives of rich NeXML files.

@@

<!-- Emphasize:

1) storing character data and tree data together, convenient for the comparative methods community
2) Adding metadata, including automated addition of metadata and semantic utilities 
3) Extending the NeXML standard  
4) Accessing / opening up existing rich NeXML data to the R user

```{r compile-settings, include=FALSE}
library(knitr)
opts_chunk$set(tidy = FALSE, warning = FALSE, message = FALSE, 
               cache = 1, comment = NA, verbose = TRUE)
basename <- gsub(".Rmd", "", knitr:::knit_concord$get('infile') )
opts_chunk$set(fig.path = paste("figure/", basename, "-", sep=""),
               cache.path = paste("cache/", basename, "/", sep=""))

```

Though R provides the most extensive and widely used collection of
comparative phylogenetic methods tools, R users lack a convenient,
extensible format to exchange and archive comparative methods data.

## Background

### Challenges with existing formats
See [Vos _et al._ 2012].

### Challenges with R formats

- Most R users rely on the NEXUS inherits the limitations of NEXUS file format

<!--
A summar of some phylogenetic formats in R:

| Class     | R package   | R Object 
|-----------|-------------|-------
| phylo     | ape         | S3 
| phylo4    | phylobase   | S4
| phylo4d   | adephylo    | S4
| ouchtree  | ouch        | S4 
| treeshape | apTreeshape | list 
-->

### NeXML

See [Vos _et al._ 2012].

Mention of data sharing challenges and opportunities.  See [Drew et al 2013]

Why RNeXML?

* Why NeXML -- because the NEXUS standard isn't a standard.  Developers extend
it arbitrarily to meet their needs, yielding mutually incompatible dialects. 
For example, NEXUS cannot represent horizontal gene transfer, ambiguous
use of symbols, cannot be extended reliably, and there is no way to
validate that it is proper NEXUS.  NeXML addresses all of these problems
[Vos _et al._ 2012] with a state-of-the-art data management format.

* Why RNeXML: Huge ecosystem of phylogenetics packages. (1) Read NeXML.
All would benefit by being able to read NeXML, both as a robust standard
free of the pitfalls of NEXUS, and more importantly, as a much richer
source of data thanks to its extendable annotation model.  (2) Write
NeXML.  R packages and R users add rich information to the phylogenies
they work with, but need a platform independent data exchange format to
share and archive this. In particular, we illustrate how NeXML can
be used to distribute both phylogenies and character trait data used in 
many popular phylogenetic comparative methods tools in R.  (3) Annotate 
and extend nexml. `RNeXML` provides convenient tools for adding metadata
to NeXML files, including the automated generation of metadata and examples
of how to extend the NeXML standard to represent new data features without
breaking validation or backwards compatability of the file type.

## Introduction to RNeXML

@@

The current stable version of the `RNeXML` package can be installed from
the CRAN library using the standard installation method <!-- Actually not yet -->

```r
install.packages("RNeXML")
```

The latest stable development version can be installed from Github using the `devtools` package:

```r
library(devtools)
install_github("RNeXML", "ropensci")
```

## Working with phylogenetic trees

#### Read methods

@@

We begin by reading in an example NeXML file provided with the package.

@@

(For convenience `nexml_read` is also aliased as `read.nexml`, for
users familiar with the somewhat analogous `read.nexus` function).
The resulting `nex` object is an R object representation of the NeXML.
From this object we can extract any phylogenies it contains in the
`ape::phylo` format:

@@

We can then leverage the rich suite of methods available from the R
phylogenetics community (including coercion into other representations,
such as `ouch` [@ouch] and `phylobase` [@phylobase] phylogeny
formats). For instance, here we use the standard plotting function from
the `ape` package to plot the resulting `ape::phylo` object returned by
`get_trees`.

```{r eval=FALSE}
plot(phy)
```

@@

Previously, this was possible using the `treebase` package [@treebase]
for R, which relied on the NEXUS parser provided in the `ape` package.
Due to inconsistencies present in the NEXUS standard [Vos _et al._
2012], a significant number <!-- State explicitly --> of TreeBASE NEXUS
files could not be read in this way.  Using the more precisely defined
NEXML standard, every valid NeXML file on TreeBASE can now successfully
be parsed to R objects.

#### Character data

@@

#### Write methods

@@

```{r}
library(geiger)
data(geospiza)
nexml_write(trees = geospiza$phy, characters = geospiza$dat, file="geospiza.xml")
```

@@

This allows us to include multiple trees and character matrices to a single NeXML file.  
<!-- We can add relevant metadata to each tree, such as the citation information 
     indicating the paper in which it was first described.  -->

### Metadata

At the heart of the NeXML standard is the ability to encode arbitrary
metadata in precisely specified, machine-readable formats that maintain
backwards compabitibilty with any existing NeXML parser. NeXML uses the
"Resource Description Framework in Annotations", or RDFa [W3C 2013],
as described in [Vos et al 2011].  These annotations can be used to add 
information to any particular element of the data, (a tree, node, edge,
or taxonomic unit; a character matrix or specific trait, or even another 
metadata annotation or the NeXML data file as a whole). `RNeXML` provides
a range of tools for accessing metadata embedded in a NeXML file.

### Writing NeXML metadata

Researchers cannot take advantage of NeXML metadata manipulation if
there is no metadata there.  The burden of adding good metadata remains a
significant barrier to increased sharing and reuse of scientific research
[Tenopir et al. 2011].  `RNeXML` seeks to reduce the burden of adding
high quality, machine-readable metadata to NeXML files. This occurs on
several levels:

- Automatic annotation.

`RNeXML` automatically adds basic metadata by default such as a timestamp
of when the tree was produced and an open data license (CC0, compatible
with deposition on Dryad data repository), which can be configured
as needed.  To adjust these and other common metadata fields that should 
be part of any data deposition, `RNeXML` provides the `add_basic_meta()`
function.  This function does not assume any familiarity with the concepts
of namespaces or other more advanced metadata tools discussed later.  The
`add_basic_meta` takes an existing `nexml` object (or will create a new one
if none is provided, like the other `add_` functions), and any of the following
fields: `title`, `description`, `creator`, `pubdate`, `rights`, `publisher`,
`citation`.  Additional fields may be added to subsequent versions, see
`?add_basic_metadata` for the most up-to-date information.  Here we create
an new `nexml` object with the basic metadata for the `bird.orders` data
provided in the `ape` package:

@@

or to papers by simply using the object's DOI:

```{r}
library("knitcitations")
geiger_nex <- add_basic_meta(geiger_nex, citation=cite("10.2307/2408428"))
```

@@

`RNeXML` also provides various functions to assist in automatically adding
other more specific metadata to various elements of the NeXML data.

For instance, misspellings or inconsistent use of taxonomic names in phylogenetic
data are a common source of frustration in reusing phylogenetic trees. By 
associating the taxonomic labels against a specific, centralized database
we can catch typographical errors and indicate more precisely which classification
we are appealing to in the event of conflicting taxonomic definitions.  In this 
example, we indicate that we intend the bird orders listed in the tree to correspond
to the NCBI definitions

@@

This function uses the `taxize` R library ([Chamberlain 2013]) to check each taxon
label against the NCBI database.  If a unique match is found, a metadata annotation
is added to the taxon providing the NCBI identification number to the taxonomic unit.
If no match is found, the user is warned to check for possible typographic errors in
the taxonomic labels provided.  <!-- what if multiple NCBI taxon IDs are found, e.g. 'Aotus'? -->

### Custom metadata extensions

So far our examples have relied on functions that have a built-in understanding of the 
kind of metadata being provided, and can thus generate the appropriate metadata with 
little further guidence.  In order to be machine-readable and understandable, 
NeXML requires all metadata annotations follow the RDFa format which explicitly declares 
the meaning of the annotation in a way a computer can understand.  For instance, the 
title we gave to the `bird.orders` phylogeny appears in the NeXML as:

@@

(The `add_namespaces` function will avoid creating any duplicate definitions)
For finer control of the level at which a `meta` element is added, we will manipulate 
the nexml R object directly using S4 subsetting, as shown later.

<!-- Ugh. y'all have much more practice explaining this well and without tangents, 
so please feel free to edit or re-write.  -->

@@

<!--
- Utilities to add common metadata

  - Add a citation to any papers associated with the phylogeny data.  
  - Writing in methodological information on how tree has been generated or analyzed (including R code)
  - Add annotations to any taxon.

Temporal, spatial, taxonomic coverage for dryad/knb?
-->

### Reading NeXML metadata

A call to the nexml object prints some metadata summarizing the data structure: <!-- need to make this method faster!-->

@@

```{r}
library(XML)
xpathSApply(rdf, "//dc:title", xmlValue) 
```

XPath is an expressive syntax for querying XML-structured data such
as RDF.  Queries using the XPath syntax are possible in R using the
XML library [Temple Lang 2013, XML].  (Note that XPath is far more
expressive for querying the metadata once the RDFa has been extracted,
since attribute values are now interpreted as namespaced XML nodes and
not just arbitrary text.)

The most powerful way to take advantage of the machine-readable nature
of the metadata involves the construction of SPARQL queries. This
approach can exploit the reasoning and logical deduction capacity of
formal vocabularies (ontologies). For a proper introduction to SPARQL
queries, see CITE SOMETHING.  Meanwhile, we just introduce the sytnax
necessary to apply such queries to NeXML files from R.  First we load
the necessary R library and import the RDF-extracted metadata:

```{r}
library(rrdf)
saveXML(rdf, "rdf_meta.xml") # rrdf requires a file name, so we must write the XML out first
lib <- load.rdf("rdf_meta.xml")
```

@@

### Extending the NeXML standard through metadata annotation.

@@

Here we illustrate this process using the example of stochastic character
mapping [Huelsenbeck et al 2003]. A stochastic character map is simply
an annotation of the branches on a phylogeny, assigning each section
of each branch to a particular "state" (typically of a morphological
characteristic).

[Bollback 2006] provides a widely used stand-alone software implementation
of this method in the software `simmap`, which modified the standard
Newick tree format to express this additional information. This can
break compatibility with other software [^1], and creates a format that
cannot be interpreted without additional information describing this
convention.  By contrast, the NeXML extension is not only backwards
compatible but contains a precise and machine-readable description of
what it is encoding.

@@

[Revell 2011] describes the `phytools` package for R, which includes
utilities for reading, manipulating, and writing `simmap` files in R.
In this example, we also show how to define `RNeXML` functions that 
map the R representation used by Revell (an extension of the `ape` class)
into the NeXML extension we have defined by using `RNeXML` functions..

@@

We can then use various functions from `phytools` designed for simmap
objects, such as the plotting function:

```{r caption="Stochastic character mapping on a phylogeny, as generated by the phytools package after parsing the simmap-extended NeXML"}
library(phytools)
plotSimmap(phy)
```

@@

[Revell 2011]: http://doi.org/10.1111/j.2041-210X.2011.00169.x
[Huelsenbeck et al 2003]: http://doi.org/10.1080/10635150390192780 "Stochastic Mapping of Morphological Characters"
[Bollback 2006]: http://doi.org/10.1186/1471-2105-7-88 "SIMMAP: stochastic character mapping of discrete traits on phylogenies."

[^1]: By using the commenting mechanism of the Newick format, it is
possible that other software that doesn't also use the comment mechanism
for some other such purpose would be able to successfully parse the tree.
However there is no way to guarentee that this is the case or for the
data format to describe its use.

### Subsetting nexml objects directly

@@

### Publishing NeXML files from R

Data archiving is increasingly required by journals in evolutionary biology and 
biodiversity (e.g. [Whitlock2011]).  The burden of preparing and submitting properly
annotated data to archives continues to be a significant barrier ([Tenopir et al. 2011], [Stodden 2010]),
and many phylogenetic trees upon which studies are based are inaccessible or lost 
to the research community [Drew et al 2013].  `RNeXML` seeks to lower these barriers
by providing functions to immediately archive NeXML files in scientific repositories.

@@

### Implementation and architecture

The `RNeXML` package is built for the NeXML Schema, [version 2009],
using the R software environment and the XML package for parsing and
generating XML [Temple Lang 2013, XML].

<!--  Find more consise way to make sure dependencies are cited. 
      How about citing packages that are only on the suggests list?

and provides conversions into
the popular `phylo` format provided by the `ape` package [Paradis 2004].
`RNeXML` can automatically check taxonomic names and add taxonomic
identifiers when generating NeXML using methods from the `taxize` package
([Chamberlain 2013]).
-->

<!-- Possibly cite all dependencies and suggests from the DESCRIPTION --> 
<!-- Reference any installation and configuration instructions necessary (e.g. API credentials for figshare) -->

### Quality control

The `RNeXML` package performs severak levels of validation on NeXML files
written from or read into R: validating the XML, and validating against
the current NeXML schema, and performing additional checks through the 
NeXML online validation tool for elements that cannot be enforced by
schema validation alone.  Validation serves to improve consistency in
data encoding which facilitates the use of NeXML across a wide variety
of platforms and software.  A lack of a consistent standard and validation
mechanism has continually plagued the current NEXUS standard and made 
interoperability difficult [Vos et al. 2011].

Several mechanisms are in place to evaluate the quality of the `RNeXML`
package functions themselves.  In addition to the examples from the
documentation and the automated package checks provided for the R system,
the `RNeXML` package includes a complete test suite using the `testthat`
package [Wickham 2011]. Package development on Github includes continuous 
integration through the `travis` platform, in which the full test suite
is run every time the code is changed.  This not only gives users confidence
that the development copy can be successfully installed, but should assist
other developers seeking to contribute to the package to test that their proposed 
changes have not negatively impacted any of the existing code base.

### Software reuse and support

The `RNeXML` package is available through the CRAN repository under
a CC0 license. At the time of writing, the current version is `r packageDescription("RNeXML", fields="Version")`. 
The current development
branch can be found on Github at [https://github.com/ropensci/RNeXML].
See the NEWS file for changes in recent versions. Users are encouraged
to submit bug reports or feature requests in the issues log.  Contact
ropensci-discuss@googlegroups.com or the package maintainer for help.

There are many promising directions for further development of the
`RNeXML` package.  The project's [Milestones page][milestones] page provides a list of
steps currently planned or under consideration.  In particular, these
milestones include increased support for interactive mode of entering data
(wizards), and support for integration of semantics defined through the
Web Ontology language (OWL), as discussed in the Challenges section.

### Acknowledgements

This project was supported in part by a grant from the Alfred P Sloan
Foundation (CB and SC), NSF Grant DBI-1306697 (CB), ... `RNeXML` is part
of the rOpenSci project, [http://ropensci.org].

>>>>

---
layout: preprint
#layout: review, 11pt
title: "RNeXML: Parsing and Serializing the Next Generation of Phyloinformatic Data in R"
author: 
  - name: Carl Boettiger
    affiliation: cstar
    email: cboettig(at)gmail.com
    footnote: Corresponding author
  - name: Hilmar Lapp
    affiliation: NESCent
  - name: Scott Chamberlain
    affiliation: SimonFraser
  - name: Rutger Vos
    affiliation: NBC
address: 
  - code: cstar
    address: | 
      Center for Stock Assessment Research, 
      Department of Applied Math and Statistics, 
      University of California, Mail Stop SOE-2,
      Santa Cruz, CA 95064, USA
  - code: NESCent
    address: | 
      National Evolutionary Synthesis Center, Durham, NC, USA
  - code: SimonFraser
    address: |
      Department of Biology, Simon Fraser University, Burnaby, Canada
  - code: NBC
    address: | 
      Naturalis Biodiversity Center, Leiden, the Netherlands
abstract: | 
      1. NeXML promises to be the next generation of phylogenetic
      informatics data and metadata exchange formats, offering substantial
      improvements in reliability, extensibility, and richness over
      earlier standards such as NEXUS. In this paper we present the RNeXML
      package which provides a bridge between this rich and reliable
      data format and the extensive library of phylogenetic tools already
      available in R, with a particular emphasis on comparative methods.

      2. RNeXML is much richer than a basic parser: here we illustrate
      how the package facilitates adding and manipulating semantic 
      annotations.

@@

      4. RNeXML is well suited for users to preserve and maintain their
      personal archives of phylogenetic and comparative trait data in consistent,
      query-able, and cross-platform format, or leverage the rapidly growing 
      public community archives of rich NeXML files.

bibliography: components/references.bib
csl: components/methods-in-ecology-and-evolution.csl
documentclass: components/elsarticle

output: 
  pdf_document:
    template: components/elsarticle.latex
    keep_tex: false
    fig_caption: true

@@

<!------------------------------------------------

Map/Outline:
------------

- Introduction
  - Exciting times in phylogenetics!  
  - Challenges with existing formats 
  - Overview of NeXML
  - Map

- Tutorial
  - Reading Phylogenetic trees
  - Reading Character data
  - Writing phylogenetic trees & character data

  - Metadata
    - Adding basic metadata
    - Adding taxonomic identifers
    - Adding Generic metadata
    - Parsing metadata. XML. RDF.
    - Extending NeXML with metadata: Example defining simmap stochastic character mappings.

  - Publishing NeXML

- Implementation, Quality Control
- Software reuse and support

------------------------------------------------>

```{r compile-settings, include=FALSE}
library("methods")
library("knitr")
opts_chunk$set(tidy = FALSE, warning = FALSE, message = FALSE, 
               cache = 1, comment = NA, verbose = TRUE)
basename <- gsub(".Rmd", "", knitr:::knit_concord$get('infile')) 
opts_chunk$set(fig.path = paste("components/figure/", basename, "-", sep=""),
               cache.path = paste("components/cache/", basename, "/", sep=""))

library("knitcitations")
cleanbib()
cite_options(citation_format = "pandoc", check.entries=FALSE)
```

# Introduction

<!-- Need to say some of this stuff without repeating too much of existing literature -->

Though R provides the most extensive and widely used collection of
comparative phylogenetic methods tools, R users lack a convenient,
extensible format to exchange and archive comparative methods data.
A common file format for sharing phylogenetic tree data has long played
an essential role in the advancement of phylogenetics research.
The rapid increase in phylogenetics data and research methods that make
use of such data has introduced many new challenges and opportunities 
to sharing this information both between software environments and 
between research groups `r citet("Drew PLOS Biology Tree of Life 2013")`.

<!-- FIXME -->

ADD STUFF ABOUT:

- Rapid growth in phylogenetics data (+citations)
- Rapid growth of research methods and software that use phylogenetic data. Comparative methods in evolution. Ecological phylogenetics. Biodiversity informatics. (+citations)
- More on data sharing challenges and opportunities a la @Drew_2013

The ever expanding needs of the phylogenetics community are rapidly
out-pacing the capacity of current standards, which is resulting in a
radiation of different data representations and a loss of interoperability
at the very time when the growth of available data and mehtods has made
that interoprability most valuable.

For example, the currently popular NEXUS file format cannot consistently
represent horizontal gene transfer or ambiguity in reading a character
(such as a DNA sequence base pair).  This has led developers to
repeatedly add small modifications to the standard to meet the needs
of their own software, which results in mutually incompatible dialects
that largely defeat the advantages of a shared standard file format.
NeXML was developed to addresses each of these problems `r citep("Vos NeXML Systematic Biology")`  
by drawing on state-of-the-art data management technology.

<!-- What do we need to say, if anything, about the current R formats?  about phylobase?  -->

The R lanugage `r citep(citation())` has become an important enviroment for 
phylogenetics research, with over 30 packages available including several
with over 1000 citations `r citep("http://cran.r-project.org/web/views/Phylogenetics.html")`,
though stand-alone programs remain an important part of this ecosystem
for tasks such as generating phylogenetic trees from sequence data.  
Within this ecosystem a variety of different data formats for
phylogenetic data have been developed and modified. Most but not all 
are based on the simple and popular `phylo` format provided by the
ape package `r citep(citation("ape"))`.

Though certain functions allow R users to switch between these formats
within R, these formats frequently contain more information than their
NEXUS file counterparts, and users lack a suitable alternative for
archiving and sharing phylogenetic data outside of the R ecosystem. 
Consequently, information added to phylogenies by these R programs cannot 
easily be shared, archived, and made available to other software;
but inherits the limiations of the NEXUS format.

`RNeXML` opens up a rich world of both currently available and rapidly
growing data sets in the NeXML format to the R user and the large ecosystem
of R packages currently avaliable for handling phylogenetic data  
R packages and R users frequently add rich information to the phylogenies
they work with, but need a platform independent data exchange format to
share and archive this. We illustrate how NeXML can
be used to distribute both phylogenies and character trait data used in 
many popular phylogenetic comparative methods tools in R.  
`RNeXML` also provides convenient tools for adding metadata
to NeXML files, including the automated generation of metadata and examples
of how to extend the NeXML standard to represent new data features without
breaking validation or backwards compatability of the file type.

<!-- FIXME something about how we will increasingly want to work with (semantic) metadata
and how RNeXML provides a solid platform from which to do so.  
-->

<!-- Map paragraph -->

In this paper, we introduce the reader to the concepts and use of
RNeXML and related technologies. We begin by discussing methods to read
phylogenetic tree data and character trait data in the NeXML format into
R, and also how to write phylogenetic data and character trait data into
NeXML. The heart of this section focuses on tools for reading and writing
the kind of semantically rich metadata that most sets NeXML apart from
other phylogenetic data formats. Starting simple, we use examples of
increasing complexity to illustrate concepts and tools from biodiversity
informatics that allow users to express and intepret complicated data
structures automatically.  We end with an example that uses metadata to
extend the NeXML standard to describe stochastic character mapping, one
of the kinds of data annotation that is currently available in R formats
but lacks a consistent, extensible export format. We illustrate how NeXML
files can be formally published to a data respository directly from R.
Finally, we provide an brief description of how the software has been
implemented, what quality control measures are in place, as well as
information about software reuse and getting support.

This paper has been written as a dynamic document 
`r citep(citation("knitr")[[2]])`.  The reader may paste the code blocks
shown into their R terminal to follow along with the exmples shown.

# Introduction to RNeXML

@@

## Getting Started

The current stable version of the `RNeXML` package can be installed from
the CRAN library[^1].

<!-- actually not yet, but before submission -->

```r
install.packages("RNeXML", dependencies=c("DEPENDS", "IMPORTS", "SUGGESTS"))
```

<!-- 
The latest stable development version can be installed from Github using the `devtools` package:

```r
library(devtools)
install_github("RNeXML", "ropensci", dependencies=c("DEPENDS", "IMPORTS", "SUGGESTS"))
```
-->

## Reading phylogenetic trees

@@

We begin by reading in an example NeXML file provided with the package.[^2]

@@

The resulting `nex` object is an R object representation of the NeXML.  
This is a native R object for representing phylogenetic trees (as well as 
character traits, and metadata), analgous to the `phylo` object from the 
popular `ape` R package `r citep(citation("ape"))`.

While R functions and packages could be written to use this `nexml` object
directly, most users will prefer to

From this object we can extract any phylogenies it contains in the
`ape::phylo` format:

@@

We can then leverage the rich suite of methods available from the R
phylogenetics community (including coercion into other representations,
such as `ouch` `r citep(citation("ouch"))` and `phylobase` `r citep(citation("phylobase"))` phylogeny
formats). For instance, here we use the standard plotting function from
the `ape` package to plot the resulting `ape::phylo` object returned by
`get_trees`.

```{r eval=FALSE}
plot(phy[[1]])
```

@@

Previously, this was possible using the `treebase` package `r citep("Boettiger Temple Lang treebase 2012")`.
for R, which relied on the NEXUS parser provided in the `ape` package.
Due to inconsistencies present in the NEXUS standard `r citep("Vos NeXML 2012")`, 
a significant number <!-- State explicitly --> of TreeBASE NEXUS
files could not be read in this way.  Using the more precisely defined
NEXML standard, every valid NeXML file on TreeBASE can now successfully
be parsed to R objects.

## Reading character data

@@

## Writing phylogenetic trees

@@

```{r}
library("geiger")
data(geospiza)
nexml_write(trees = geospiza$phy, characters = geospiza$dat, file="geospiza.xml")
```

These examples use data from the popular phylogenetics software `geiger` `r citep(citation("geiger"))`.

@@

This allows us to include multiple trees and character matrices to a single NeXML file.

# Metadata

At the heart of the NeXML standard is the ability to encode arbitrary
metadata in precisely specified, machine-readable formats that maintain
backwards compabitibilty with any existing NeXML parser. NeXML uses the
"Resource Description Framework in Annotations", or RDFa `r citep("http://www.w3.org/TR/xhtml-rdfa-primer")`,
as described in @Vos_2012.  These annotations can be used to add 
information to any particular element of the data, (a tree, node, edge,
or taxonomic unit; a character matrix or specific trait, or even another 
metadata annotation or the NeXML data file as a whole). `RNeXML` provides
a range of tools for accessing metadata embedded in a NeXML file.

## Writing NeXML metadata

Researchers cannot take advantage of NeXML metadata manipulation if
there is no metadata there.  The burden of adding good metadata remains a
significant barrier to increased sharing and reuse of scientific research
`r citep("Tenopir Data Sharing by Scientists: Practices and Perceptions")`.
`RNeXML` seeks to reduce the burden of adding
high quality, machine-readable metadata to NeXML files. This occurs on
several levels:

- Automatic annotation
- Adding common metadata fields (`add_basic_metadata`)
- Taxonomic identifiers (`taxize_nexml`)
- Generic metadata (`add_meta`)
- Extending the NeXML standard using RDFa

`RNeXML` automatically adds basic metadata by default such as a timestamp
of when the tree was produced and an open data license (CC0, compatible
with deposition on Dryad data repository), which can be configured
as needed.

To adjust these and other common metadata fields that should 
be part of any data deposition, `RNeXML` provides the `add_basic_meta()`
function.  This function does not assume any familiarity with the concepts
of namespaces or other more advanced metadata tools discussed later.  The
`add_basic_meta` takes an existing `nexml` object (or will create a new one
if none is provided, like the other `add_` functions), and any of the following
fields: `title`, `description`, `creator`, `pubdate`, `rights`, `publisher`,
`citation`.  Additional fields may be added to subsequent versions, see
`?add_basic_metadata` for the most up-to-date information.  Here we create
an new `nexml` object with the basic metadata for the `bird.orders` data
provided in the `ape` package:

@@

or to papers by simply using the object's DOI (using `r citet(citation("knitcitations"))`):

```{r}
library("knitcitations")
geiger_nex <- add_basic_meta(geiger_nex, citation = bib_metadata("10.2307/2408428"))
```

@@

## Taxonomic identifiers

`RNeXML` also provides various functions to assist in automatically adding
other more specific metadata to various elements of the NeXML data.
One common use case is that of taxonomic identifers.  Misspellings or
inconsistent use of taxonomic names in phylogenetic data are a common
source of frustration in reusing phylogenetic trees. By associating the
taxonomic labels against a specific, centralized database we can catch
typographical errors and indicate more precisely which classification
we are appealing to in the event of conflicting taxonomic definitions.
In this example, we indicate that we intend the bird orders listed in
the tree to correspond to the NCBI definitions

@@

This function uses the `taxize` R library `r citep("Chamberlain taxize F1000Research 2013")` to check each taxon label against the NCBI database.
If a unique match is found, a metadata annotation is added to the taxon
providing the NCBI identification number to the taxonomic unit.  If no
match is found, the user is warned to check for possible typographic
errors in the taxonomic labels provided.

<!-- what if multiple NCBI taxon IDs are found, e.g. 'Aotus'? -->

## Custom metadata extensions

So far our examples have relied on functions that have a built-in
understanding of the kind of metadata being provided, and can thus
generate the appropriate metadata with little further guidence.
In order to be machine-readable and understandable, NeXML requires
all metadata annotations follow the RDFa format (see @Vos_2012) which
explicitly declares the meaning of the annotation in a way a computer
can understand.  For instance, the title we gave to the `bird.orders`
phylogeny appears in the NeXML as:

@@

Note that the `add_namespaces` function will automatically avoid creating
any duplicate definitions.

For finer control of the level at which a `meta` element is added,
we will manipulate the nexml R object directly using S4 subsetting,
as shown later.

<!--FIXME Ugh. y'all have much more practice explaining this well and without tangents, 
so please feel free to edit or re-write.  -->

@@

## Reading NeXML metadata

A call to the nexml object prints some metadata summarizing the data structure:

@@

```{r}
library("XML")
xpathSApply(rdf, "//dc:title", xmlValue) 
```

XPath is an expressive syntax for querying XML-structured data such
as RDF.  Queries using the XPath syntax are possible in R using the
XML library `r citep(citation("XML"))`.  (Note that XPath is far more
expressive for querying the metadata once the RDFa has been extracted,
since attribute values are now interpreted as namespaced XML nodes and
not just arbitrary text.)

The most powerful way to take advantage of the machine-readable nature
of the metadata involves the construction of SPARQL queries. This
approach can exploit the reasoning and logical deduction capacity of
formal vocabularies (ontologies). For a proper introduction to SPARQL
queries, see `r citet("Hartig Introduction to SPARQL")`.  Meanwhile, we just introduce the sytnax
necessary to apply such queries to NeXML files from R.  First we load
the necessary R library `r citep(citation("rrdf"))` and import the RDF-extracted metadata:

```{r}
library("rrdf")
saveXML(rdf, "rdf_meta.xml") # rrdf requires a file name, so we must write the XML out first
lib <- load.rdf("rdf_meta.xml")
```

@@

## Extending the NeXML standard through metadata annotation.

@@

Here we illustrate this process using the example of stochastic character
mapping `r citep("Huelsenbeck stochastic character mapping")`. A stochastic character map is simply
an annotation of the branches on a phylogeny, assigning each section
of each branch to a particular "state" (typically of a morphological
characteristic).

`r citet("Bollback simmap 2006")` provides a widely used stand-alone software implementation
of this method in the software `simmap`, which modified the standard
Newick tree format to express this additional information. This can
break compatibility with other software [^3], and creates a format that
cannot be interpreted without additional information describing this
convention.  By contrast, the NeXML extension is not only backwards
compatible but contains a precise and machine-readable description of
what it is encoding.

@@

`r citet("Revell phytools 2011")` describes the `phytools` package for R, which includes
utilities for reading, manipulating, and writing `simmap` files in R.
In this example, we also show how to define `RNeXML` functions that 
map the R representation used by Revell (an extension of the `ape` class)
into the NeXML extension we have defined by using `RNeXML` functions.

@@

We can then use various functions from `phytools` designed for simmap
objects `r citep(citation("phytools"))`, such as the plotting function:

```{r caption="Stochastic character mapping on a phylogeny, as generated by the phytools package after parsing the simmap-extended NeXML"}
library("phytools")
plotSimmap(phy)
```

@@

<!--
[Revell 2011]: http://doi.org/10.1111/j.2041-210X.2011.00169.x
[Huelsenbeck et al 2003]: http://doi.org/10.1080/10635150390192780 "Stochastic Mapping of Morphological Characters"
[Bollback 2006]: http://doi.org/10.1186/1471-2105-7-88 "SIMMAP: stochastic character mapping of discrete traits on phylogenies."
-->

<!-- Consider moving to the supplement? -->

## Understanding the `nexml` S4 object

@@

## Publishing NeXML files from R

Data archiving is increasingly required by journals in evolutionary biology and 
biodiversity (e.g. `r citet("Whitlock data archiving Evolution 2010")`).  The burden of preparing and submitting properly
annotated data to archives continues to be a significant barrier (@Tenopir_2011, `r citet("10.2139/ssrn.1550193")`),
and many phylogenetic trees upon which studies are based are inaccessible or lost 
to the research community [@Drew_2013].  `RNeXML` seeks to lower these barriers
by providing functions to immediately archive NeXML files in scientific repositories.

@@

# Implementation, architecture and quality control

The `RNeXML` package is built for the NeXML Schema, [version 2009],
using the R software environment and the XML package for parsing and
generating XML `r citep(citation("XML"))`,  provides conversions into
the popular `phylo` format provided by the `ape` package `r citep(citation("ape"))`
`RNeXML` can automatically check taxonomic names and add taxonomic
identifiers when generating NeXML using methods from the `taxize` package `r citep(citation("taxize"))`.
Unique identifiers are provided with the `uuid` package `r citep(citation("uuid"))`,
and internet interfaces by the `r citep(citation("httr"))` package. The
packages `plyr` and `reshape2``r citep(c(citation("plyr"), citation("reshape2")))` are also imported for internal methods .

<!--  Find more consise way to make sure dependencies are cited. 
      How about citing packages that are only on the suggests list? --> 
--> 
<!-- Possibly cite all dependencies and suggests from the DESCRIPTION --> 
<!-- Reference any installation and configuration instructions necessary (e.g. API credentials for figshare) -->

The `RNeXML` package performs several levels of validation on NeXML files
written from or read into R: validating the XML, and validating against
the current NeXML schema, and performing additional checks through the 
NeXML online validation tool for elements that cannot be enforced by
schema validation alone.  Validation serves to improve consistency in
data encoding which facilitates the use of NeXML across a wide variety
of platforms and software.  A lack of a consistent standard and validation
mechanism has continually plagued the current NEXUS standard and made 
interoperability difficult [@Vos_2012].

Several mechanisms are in place to evaluate the quality of the `RNeXML`
package functions themselves.  In addition to the examples from the
documentation and the automated package checks provided for the R system,
the `RNeXML` package includes a complete test suite using the `testthat`
package `r citep(citation("testthat"))`. Package development on Github includes continuous 
integration through the `travis` platform, in which the full test suite
is run every time the code is changed.  This not only gives users confidence
that the development copy can be successfully installed, but should assist
other developers seeking to contribute to the package to test that their proposed 
changes have not negatively impacted any of the existing code base.

# Software reuse and support

The `RNeXML` package is available through the CRAN repository under
a CC0 license. At the time of writing, the current version is `r packageDescription("RNeXML", fields="Version")`. 
The current development
branch can be found on Github at [https://github.com/ropensci/RNeXML].
See the NEWS file for changes in recent versions. Users are encouraged
to submit bug reports or feature requests in the issues log.  Contact
`ropensci-discuss@googlegroups.com` or the package maintainer for help,
or email the phylogenetics R users group list at `r-sig-phylo@r-project.org`.

There are many promising directions for further development of the
`RNeXML` package.  The project's [Milestones page][https://github.com/ropensci/RNeXML/issues/milestones] page provides a list of
steps currently planned or under consideration.  In particular, these
milestones include increased support for interactive mode of entering data
(wizards), and support for integration of semantics defined through the
Web Ontology language (OWL), as discussed in the Challenges section.

## Acknowledgements

This project was supported in part by a grant from the Alfred P Sloan
Foundation (CB and SC), NSF Grant DBI-1306697 (CB), ... <!-- FIXME Rutger, Hilmar, add whatever you'd like to acknowledge here -->
